### 2024-11-27 misc/expo/additiveout.tex
lecture nov 26

\bye

>>>>

\parskip=0pt
\hyphenpenalty=-1000 \pretolerance=-1 \tolerance=1000
\doublehyphendemerits=-100000 \finalhyphendemerits=-100000
\frenchspacing
\def\bref#1{[#1]}
\def\beginref{\noindent}
\def\endref{\medskip}
\vskip\parskip

\beginref
Antal Balog
and Endre Szemer\'edi,
``A statistical theorem of set addition,''
{\sl Combinatorica}\/
{\bf 14}
(1994),
263--268.
\endref
\beginref
William Timothy Gowers,
``A new proof of Szemer\'edi's theorem for arithmetic progressions of length four,''
{\sl Geometric and Functional Analysis}\/
{\bf 8}
(1998),
529--551.
\endref
\beginref
Askold Khovanskii,
``Newton polyhedron, Hilbert polynomial, and sums of finite sets,''
{\sl Functional Analysis and its Applications}\/
{\bf 26}
(1992),
276--281.
\endref
\beginref
Melvyn Bernard Nathanson
and Imre Zolt\'an Ruzsa,
``Polynomial growth of sumsets in abelian semigroups,''
{\sl Journal de Th\'eorie des Nombres de Bordeaux}\/
{\bf 14}
(2002),
553--560.
\endref
\beginref
Giorgis Petridis,
``New proofs of Pl\"unnecke-type estimates for product sets in groups.,''
{\sl Combinatorica}\/
{\bf 32}
(2012),
721--733.
\endref
\beginref
Helmut Pl\"unnecke,
{\sl Eigenschaften und Absch\"atzungen von Wirkungsfunktionen}
(Bonn:
Berichte der Gesellschaft f\"ur Mathematik und Datenverarbeitung,
1969).
\endref
\beginref
Klaus Friedrich Roth,
``On certain sets of integers,''
{\sl Journal of the London Mathematical Society}\/
{\bf 28}
(1953),
104--109.
\endref
\beginref
Imre Zolt\'an Ruzsa,
``An application of graph theory to additive number theory,''
{\sl Scientia, Series A}\/
{\bf 3}
(1989),
97--109.
\endref
\beginref
Imre Zolt\'an Ruzsa, ``Sums of finite sets,'' {\sl Number Theory: New York Seminar 1991--1995} (1996).
\endref
\beginref
Endre Szemer\'edi,
``On sets of integers containing no $k$ elements in arithmetic progression,''
{\sl Acta Arithmetica}\/
{\bf 27}
(1975),
199--245.
\endref
\beginref
Terence Tao
and Van Ha Vu,
{\sl Additive Combinatorics}
(Cambridge:
Cambridge University Press,
2006).
\endref
\beginref
Bartel Leendert van~der~Waerden,
``Beweis einer Baudetschen Vermutung,''
{\sl Nieuw Archief voor Wiskunde}\/
{\bf 15}
(1927),
212--216.
\endref
\beginref
\goodbreak\bye

### 2024-11-27 teaching/math240f24/draft.tex
lecture nov 26

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 18.\kern .5em Permutations and
combinations}{92}}

@@

Let $1\le k\le p-1$ and expand
$${p\choose k} = {p!\over k!(p-k)!}$$
Multiplying both sides by $k!(p-k)!$, we get
$$k!(p-k)!{p\choose k} = p! = p(p-1)!.$$
This is an integer and the
right-hand side is a multiple of $p$, so the left-hand side is as well. Then since $p$ divides
the product $k!(p-k)!{p\choose k}$, either $p$ divides $k!(p-k)!$ or it divides $p\choose k$.
But
$$k!(p-k)! = k(k-1)\cdots 2\cdot 1\cdot (p-k)(p-k-1)\cdots 2\cdot 1,$$
and all of the factors on the right-hand side are positive integers less than $p$. So $p$ cannot divide
any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug
\bye

\medskip\boldlabel\datestamp{24}{} A day spent selling milk.
The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
without necessarily expanding the whole thing out. To illustrate this, consider the following
scenario. Imagine that a farmer runs a roadside
milk stand. Each time a customer visits they either
\medskip
\item{i)} pay her $\$6$ and buy one bottle of white, chocolate, or strawberry milk; or
\smallskip
\item{ii)} sell her back a glass bottle for $\$1$.
\medskip
So after one customer visits, there are three ways in which the farmer could have
made $\$6$, and one way in which she could have made $-\$1$. Let us arrange these
possibilities as a function of $x$; we write $3x^6 + x^{-1}$.

@@

\medskip\boldlabel\datestamp{25} Subset sums.
Let $A\subseteq \{1,2,\ldots, 100\}$ with $|A|=10$. Prove
that there has to be subsets $X\subseteq A$ and $Y\subseteq A$ with $X\ne Y$ such that
$$\sum_{x\in X} x = \sum_{y\in Y} y.$$
We define $f:2^A\to \NN$ by letting
$$f(X) = \sum_{x\in X} x.$$
(So $f(\emptyset) = 0$.) What is the maximum possible sum? Well, this happens when
$A = \{100,99, 98,\ldots, 91\}$ and $X = A$, in which case
$$f(X) = \sum_{x\in X} x \le 100 + 99 + 98 + \cdots + 91 \le 1000.$$
The minimum possible sum is $0$, when $X = \emptyset$,
Hence let $2^A$ be the set of pigeons and let $\{0,1,2,\ldots,1000\}$ be the set of holes
There are $2^{|A|} = 2^{10} = 1024$ pigeons and $1001$ holes, so by the pigeonhole principle
there must be two pigeons that nest in the same hole; in other words, $f$ is not injective.
Hence there are $X,Y\in 2^A$ with $X\ne Y$ such that $f(X) = f(Y)$.

@@

Let $(a_n) = a_0, a_1, a_2, \ldots$ be a sequence of real numbers (in all
the examples we consider, they will be integers). A {\it recurrence relation}
for the sequence is a formula for $a_n$ in terms of some of the $a_m$ with $m<n$.
For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 1$. The first few terms of $F_n$ are
$$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
uses these numbers to describe the growth of a population of rabbits under certain idealised
conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
to the number different ways to arrange long and short Sanskrit syllables into lines of poetry.

>>>>

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 13.\kern .5em Triangles and bipartite
graphs}{74}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 14.\kern .5em Trees}{78}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 15.\kern .5em Eulerian
trails and circuits}{81}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 16.\kern .5em Planar graphs}{83}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx IV. Combinatorics \kern 3pt}}{88}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 17.\kern .5em Basic counting techniques}{89}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 18.\kern .5em Permutations and
combinations}{92}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 19.\kern .5em Applications of the
pigeonhole principle}{100}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 20.\kern .5em Recurrences}{102}}

@@

Let $1\le k\le p-1$ and expand
$${p\choose k} = {p!\over k!(p-k)!}$$
Multiplying both sides by $k!(p-k)!$, we get
$$k!(p-k)!{p\choose k} = p! = p(p-1)!.$$
This is an integer and the
right-hand side is a multiple of $p$, so the left-hand side is as well. Then since $p$ divides
the product $k!(p-k)!{p\choose k}$, either $p$ divides $k!(p-k)!$ or it divides $p\choose k$.
But
$$k!(p-k)! = k(k-1)\cdots 2\cdot 1\cdot (p-k)(p-k-1)\cdots 2\cdot 1,$$
and all of the factors on the right-hand side are positive integers less than $p$. So $p$ cannot divide
any of them, and hence $p$ does not divide $k!(p-k)!$. We conclude that $p\divides {p\choose k}$.\slug

\medskip\boldlabel\datestamp{26}{xi} A day spent selling milk.
The binomial theorem can be used to find the value of a specific coefficient in a polynomial,
without necessarily expanding the whole thing out. To illustrate this, consider the following
scenario. Imagine that a farmer runs a roadside
milk stand. Each time a customer visits they either
\medskip
\item{i)} pay her $\$6$ and buy one bottle of white, chocolate, or strawberry milk; or
\smallskip
\item{ii)} sell her back a glass bottle for $\$1$.
\medskip
So after one customer visits, there are three ways in which the farmer could have
made $\$6$, and one way in which she could have made $-\$1$. Let us arrange these
possibilities as a function of $x$; we write $3x^6 + x^{-1}$.

@@

\medskip\boldlabel Subset sums.
Let $A\subseteq \{1,2,\ldots, 100\}$ with $|A|=10$. Prove
that there has to be subsets $X\subseteq A$ and $Y\subseteq A$ with $X\ne Y$ such that
$$\sum_{x\in X} x = \sum_{y\in Y} y.$$
We define $f:2^A\to \NN$ by letting
$$f(X) = \sum_{x\in X} x.$$
(So $f(\emptyset) = 0$.) What is the maximum possible sum? Well, this happens when
$A = \{100,99, 98,\ldots, 91\}$ and $X = A$, in which case
$$f(X) = \sum_{x\in X} x \le 100 + 99 + 98 + \cdots + 91 \le 1000.$$
The minimum possible sum is $0$, when $X = \emptyset$,
Hence let $2^A$ be the set of pigeons and let $\{0,1,2,\ldots,1000\}$ be the set of holes
There are $2^{|A|} = 2^{10} = 1024$ pigeons and $1001$ holes, so by the pigeonhole principle
there must be two pigeons that nest in the same hole; in other words, $f$ is not injective.
Hence there are $X,Y\in 2^A$ with $X\ne Y$ such that $f(X) = f(Y)$.

@@

Let $(a_n) = a_0, a_1, a_2, \ldots$ be a sequence of real numbers (in all
the examples we consider, they will be integers). A {\it recurrence relation}
for the sequence is a formula for $a_n$ in terms of some of the $a_m$ with $m<n$.
For instance, the {\it Fibonacci sequence} is the sequence $F_n$ defined by $F_0 = 0$,
$F_1 = 1$, and $F_n = F_{n-1} + F_{n-2}$ for $n\ge 1$. The first few terms of $F_n$ are
$$ 0 , 1, 1, 2, 3, 5, 8, 13, 21, \ldots.$$
The sequence is named for the Italian mathematician Fibonacci, who in his 1202 book {\sl Liber Abaci}
uses these numbers to describe the growth of a population of rabbits under certain idealised
conditions. The sequence was known to Indian linguists many centuries earlier, as it pertains
to the number of different ways to arrange long and short Sanskrit syllables into lines of poetry.
\bye
